---
title: "Picrust"
author: "SD"
date: "2026-03-23"
output: html_document
---

```{r library, include=FALSE}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(ggrepel)
```
# Picrust Analysis of MetaCyc
This pipeline will be a complex analysis of the Picrust output we did on the class server.

## Input and Convert the Needed Files

Input and convert the ```phyloseq_taxonomy.rds``` sample data into a data frame and move the ```fastq``` sample IDs into a column. Then import the ```path_abun_unstrat.tsv``` file.

```{r Import Data}
meta = readRDS('../datasets/ps_filtered.rds') %>% .@sam_data %>% data.frame() %>% 
  rownames_to_column('sample_name')

metacyc = read_delim('../datasets/picrust_output/pathways_out/path_abun_unstrat.tsv')
```

```{r Group}
fast = c('accelerator','crusier sprinter','manoeuvrer')
meta = meta %>%
  mutate(swim_group = ifelse(swim_performance %in% fast,
                      'Fast',
                      ifelse(swim_performance == 'generalist',
                      'Moderate',
                      'Slow'
                      )
  ),
  .after = swim_performance
  )

### QUESTION IS HERE ###
# Which pairwise comparison should I use (or all three)?

meta_fast_slow = meta %>%
  filter(
    swim_group == c('Fast','Slow')
  )

meta_mod_slow = meta %>%
  filter(
    swim_group == c('Moderate','Slow')
  )

meta_fast_mod = meta %>%
  filter(
    swim_group == c('Fast','Moderate')
  )
```


## Run the Pipeline

```{r DAA}

# 1. Subset metadata to hindgut only

### QUESTION HERE ###
# Change meta to the variables above
# Which comparision should I use?
hindgut <- meta %>% filter(sample_type == "hindgut")

# 2. Keep only hindgut samples that actually exist in MetaCyc
valid_samples <- intersect(hindgut$sample_name, colnames(metacyc))


# 3. Filter MetaCyc to those samples
metacyc_filt <- metacyc %>% 
  select(pathway,all_of(valid_samples)) %>% 
  column_to_rownames("pathway")

valid_samples = metacyc [
  ,hindgut$sample_name
]

# Perform pathway differential abundance analysis (DAA) using LinDA method
daa_results_df = pathway_daa(abundance = metacyc_filt,
                             metadata = hindgut, 
                             group = "swim_group", 
                             daa_method = "LinDA", 
                             select = NULL, reference = NULL)

# saveRDS(daa_results_df, 'meta_fish_daa_results_df.rds')
```

## Convert to MetaCyc

```{r}
# Annotate pathway results using MetaCyc

daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = daa_results_df,
                                              ko_to_kegg = FALSE)

saveRDS(daa_annotated_results_df,'../datasets/fish_daa_meta_annotated_results_df.rds')

#daa_annotated_results_df<-readRDS('../datasets/fish_daa_meta_annotated_results_df.rds')
```


## View Sample Similarity

```{r Volcano Plot}
# This code is optimized for the slow fast comparison

### TO FIX ###
# Figure out ggrepel here
volcano_meta = ggplot(data = daa_annotated_results_df, aes(x=log2FoldChange, y = -log10(p_adjust)))+
  geom_point(colour = '#8E7CA6')+
  geom_text(
    aes(
      label = ifelse(
        -log10(p_adjust)>-log10(0.05),
        as.character(description),
        ''
      )
    ),
    angle = 45,
    hjust = -0.05,
    vjust = -0.5
  )+
  aes(alpha = 0.25)+
  geom_hline(yintercept = -log10(0.05), linetype = 'dashed', color = '#E07A9A')+
  ylim(NA,2.5)+
  theme(legend.position='none')
volcano_meta
ggsave('../plots/meta_volcano.jpg',
       plot = volcano_meta,
       height=8.5, width = 14)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
